Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-01-22 Manuscript/manuscript.Rmd
edits

![**Illustrating the resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Four types of associations between bins and genomic annotations (green shapes) were then used to build the feature space including _Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. ](Figures/model_construction_and_predictor_types.png)

We applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs we obtained a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was found to be both severe and invariant toward resolution. Specifically, we found the imbalance between TAD boundary and non-TAD boundary regions to be 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 2). Furthermore, the data summaries presented in Table 1 were found to be similar to those exibited by the K562 cell line (Supplementary Table 3).

@@

## Three types of predictors define the associations between TAD boundary regions and genomic annotations

A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 1). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1). Counts of overlaps (**Overlap Counts**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**Overlap Percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

## Building an ensemble framework for TAD boundary prediction

To account for the heavily imbalanced classes reported above, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using a particular predictor type *P = {OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 5-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 (= 2520)$ models. <!--Additional performance metrics are reported in the Supplementary materials.-->

>>>>

We applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Four types of associations between bins and genomic annotations (green shapes) were then used to build the feature space including _Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_.](Figures/model_construction_and_predictor_types.png)

We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was found to be both severe and invariant toward resolution. Specifically, we found the imbalance between TAD boundary and non-TAD boundary regions to be 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Furthermore, the data summaries presented in Table 1 were found to be similar to those exibited by the K562 cell line (Supplementary Table 3).

@@

Our computational framework was used to identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using a particular predictor type *P = {Signal, OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, we trained a random forest classifier on chromosome-specific sections of the genome using 5-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 \times 21 (= 3360)$ models.
